A data-driven study of physiotherapy for cystic fibrosis: identifying sub-types of airway clearance and physical activity patterns using unsupervised clustering

### 2019-09-05 draft.tex
ACT: Kmeans metrics and interpretation

\begin{table}[H]
  \caption{Steps Clusters Interpretation}
  \label{steps_inter}
  \centering
  \begin{tabular}{ c|l }
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Interpretation \\
    \hline
    XS & Waveforms with very little steps activity or high amount of missing data \\
    \hline
    S & Moderate step activity, consistent cadence with few periods of high activity\\
    \hline
    M & Consistent moderate to high step rate with a period of sustained high step activity \\
    \hline
    L & High step rate with longer periods of sustained high step activity \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{table}[H]
  \caption{Heartrate Clusters Interpretation}
  \label{hr_inter}
  \centering
  \begin{tabular}{ c|l }
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Interpretation \\
    \hline
    S & Small heart rate values and heart rate consistently below 120 \\
    \hline
    SM & Consistent variations around 100 with very few spikes in heart rate values \\
    \hline
    M1 &  Waveform with high variability and periods with spikes above 120 \\
    \hline
    M2 & Consistent variations around 100 with spikes in values above 120 \\
    \hline
    L &  Large values around and above 120 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\section{Usage of tables and figures}
\subsection{Figures}

\subsection{Tables}
\begin{table}
  \caption{Sample table title}
  \label{sample_table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \cmidrule(r){1-2}
    Col1 & Col2 & Col3 \\
    \hline
    a & 0 & z\\
    b & 1 & y \\
    c & 2 & x \\
    \bottomrule
  \end{tabular}
\end{table}

Here is how you can reference a table \ref{sample_table}

@@

\begin{longtable}{  n|d }
\toprule
\cmidrule(r){1-2}
\textbf{Feature Name} & \textbf{Feature description} \\
\hline
\endfirsthead
Minutes wear & Number of minutes in a day with heartrate value present \\
\hline
Percentage of wear & $\frac{\text{Minutes wear}}{24\cdot60}\cdot100\%$\\
\hline
Wear during sleep & Indicator value: 1,  worn during sleep or 0, otherwise \\
\hline
Active hours & Number of hours with a step count > 500  \\
\hline
Low active hours & Number of hours with a step count < 500  \\
\hline
Mean hourly steps & Average of hourly step count \\
\hline
Step count & Total step count per day \\
\hline
Normalised step count & Rate of steps per minute when fitbit is worn \\
 & $\frac{Step Count}{Minutes Worn}$ \\
\hline
 & Number of minutes with step count > 100 with the following constraint: \\
Active Minutes Step 0 & - No minimum of consecutive minutes\\
Active Minutes Step 2 & - At least 2 consecutive minutes > 100 \\
Active Minutes Step 5 & - At least 5 consecutive minutes > 100 \\
Active Minutes Step 10 & - At least 10 consecutive minutes > 100 \\
Active Minutes Step 20 & - At least 20 consecutive minutes > 100 \\
\hline
Active minutes Step for the active period & Number of active minutes with step > 100 from 6am to midnight \\
\hline
Coefficient of Variation Step & Coefficient of variation for steps (removing zero values) at day level \\
 & $CoV = \frac{Standard Deviation}{Mean} * 100$ \\
 \hline
 Normalised step count for periods of : & \\
 - moderate activity & Rate of steps per minute when the heartrate is between 120 and 140. \\ 
 - vigorous activity & Rate of steps per minute when the heartrate is above 140. \\
 - moderate and vigorous activity & Rate of steps per minute when the heartrate is above 120. \\
 - non-moderate and non-vigorous activity & Rate of steps per minute when the heartrate is below 120.\\
 \hline
 Proxy for Resting Heartrate & Mean of the lowest 5 heartrate minutes averages \\
 \hline
 & Number of minutes with heartrate > 120 with the following constraint: \\
Active Minutes Heartrate 0 & - No minimum of consecutive minutes\\
Active Minutes Heartrate 2 & - At least 2 consecutive minutes > 120 \\
Active Minutes Heartrate 5 & - At least 5 consecutive minutes > 120 \\
Active Minutes Heartrate 10 & - At least 10 consecutive minutes > 120 \\
Active Minutes Heartrate 20 & - At least 20 consecutive minutes > 120 \\
\hline
Active minutes Heartrate for the active period & Number of active minutes with heartrate > 100 from 6am to midnight \\
\hline
Coefficient of Variation Heartrate & Coefficient of variation for heartrate (removing zero values) at day level \\
 & $CoV = \frac{Standard Deviation}{Mean} * 100$ \\
 \hline
 Number of threshold switches above 100 & Number of times the heartrate goes above 100 from below this threshold at day level. \\ 
 Number of threshold switches above 120 & Number of times the heartrate goes above 120 from below this threshold at day level. \\ 
 Number of threshold switches above 140 & Number of times the heartrate goes above 140 from below this threshold at day level. \\ 
 \hline
 & Number of minutes with step count > 100 and heartrate > 120 with the following constraint: \\
Active Minutes Step and Heartrate 0 & - No minimum of consecutive minutes\\
Active Minutes Step and Heartrate 2 & - At least 2 consecutive minutes \\
Active Minutes Step and Heartrate 5 & - At least 5 consecutive minutes \\
Active Minutes Step and Heartrate 10 & - At least 10 consecutive minutes \\
Active Minutes Step and Heartrate 20 & - At least 20 consecutive minutes \\
\hline
Heartrate Hourly max, min and median & Maximum. minimum and median of average hourly heartrate. \\
\bottomrule
\end{longtable}

@@

\begin{longtable}{ n|d}
\toprule
\cmidrule(r){1-2}
\textbf{Feature Name} & \textbf{Feature description} \\
\hline
Breath Count & Number of breaths the person has made though an ACT  \\ 
 ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
 \hline
 Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
 Minimum Breath Amplitude & Similarly minimum, maximum and standard deviation of breath amplitude. \\
 Maximum Breath Amplitude & ~ \\
 Standard Deviation of Breath Amplitude & ~ \\
 \hline
 Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment. \\ 
 Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
 Maximum Break Duration & breath duration in a treatment. \\
 Standard Deviation of Breath Duration & ~ \\
 \hline
 Break Count & Number of breaks (between breaths) in a treatment. \\
 \hline
 Mean Break Duration & Average duration (in seconds) of breaks between breaths. \\ 
 Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
 Maximum Break Duration & breaks duration in a treatment. \\
 Standard Deviation of Break Duration & ~ \\
 \hline
 Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
 ~ & sets. Set is a group of breaths separated by huffing and coughing activity. \\
 \hline
 Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
 ~ & Used to filter out "test" device sessions that are not part of a treatment. \\
 \hline
 Pressures Readings Completeness Score & Data completeness score for the day \\
 ~ & If a treatment has less than 150 pressure readings it's completeness \\
 ~ & score is 0. \\
 \hline
 Treatment Duration & Treatment duration (in minutes). \\
 \hline
 Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
 ~ & given day). \\
 \hline
 Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
 ~ & out of the prescribed amount of breaths. \\ 
 \hline
 Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
 ~ & out of the prescribed number of treatments for this period. \\
 \hline
 Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days \\
 ~ & (for a given date) out of the prescribed amount of breaths for this period. \\ 
 \hline
 Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
 ~ & out of the prescribed number of treatments for this period. \\
 \hline
 Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
 ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
 \hline
 Cumulative Adherence Score & Cumulative adherence score to date. \\
 ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
 \hline
 Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
 \hline
 Skipped Days Count & Skipped days since the last treatment. \\
 \bottomrule
\end{longtable}

@@

\begin{table}[H]
  \caption{Steps clusters Metrics Kmeans}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Size & Average Silhouette width \\
    \hline
    XS & 1099 & 0.95 \\
    S & 3942 & 0.34 \\
    M & 2737 & 0.27 \\
    L & 1982 & 0.24 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{table}[H]
  \caption{Steps clusters Metrics HC}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Cluster number & Cluster Size & Average Silhouette width \\
    \hline
    1 & 4996 & 0.12 \\
    2 & 4056 & 0.11 \\
    3 & 677 & 0.18 \\
    4 & 31 & 0.41 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{table}[H]
  \caption{Heartrate clusters Metrics Kmeans}
  \label{hr_metrics}
  \centering
  \begin{tabular}{ |c|c|c|}
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Size & Average Silhouette width \\
    \hline
    S & 1246 & 0.28 \\
    SM & 1640 & 0.2 \\
    M1 & 2308 & 0.21 \\
    M2 & 2584 & 0.26 \\
    L & 1982 & 0.33 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

Calinski-harabasz index for the heartrate clusters is 135.51 .

\begin{table}[H]
  \caption{Heartrate clusters Metrics HC}
  \label{hr_metrics}
  \centering
  \begin{tabular}{ |c|c|c|}
    \toprule
    \cmidrule(r){1-2}
    Cluster number & Cluster Size & Average Silhouette width \\
    \hline
    1 & 4221 & -0.07 \\
    2 & 4901 & -0.06 \\
    3 & 607 & -0.10 \\
    4 & 24 & 0.07 \\
    5 & 7 & -0.06 \\
    \bottomrule
    \end{tabular}
\end{table}

>>>>

\begin{table}[H]
  \caption{Steps Clusters Interpretation}
  \label{steps_inter}
  \centering
  \begin{tabular}{ c|l }
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Interpretation \\
    \midrule
    XS & Waveforms with very little steps activity or high amount of missing data \\
    \midrule
    S & Moderate step activity, consistent cadence with few periods of high activity\\
    \midrule
    M & Consistent moderate to high step rate with a period of sustained high step activity \\
    \midrule
    L & High step rate with longer periods of sustained high step activity \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{table}[H]
  \caption{Heartrate Clusters Interpretation}
  \label{hr_inter}
  \centering
  \begin{tabular}{ c|l }
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Interpretation \\
    \midrule
    S & Small heart rate values and heart rate consistently below 120 \\
    \midrule
    SM & Consistent variations around 100 with very few spikes in heart rate values \\
    \midrule
    M1 &  Waveform with high variability and periods with spikes above 120 \\
    \midrule
    M2 & Consistent variations around 100 with spikes in values above 120 \\
    \midrule
    L &  Large values around and above 120 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

@@

\begin{longtable}{  n | d }
\toprule
\cmidrule(r){1-2}
\textbf{Fitbit features} & \textbf{Feature description} \\
\midrule
\endfirsthead
Minutes wear & Number of minutes in a day with heartrate value present \\
\midrule
Percentage of wear & $\frac{\textit{Minutes wear}}{24\cdot60}\cdot100\%$\\
\midrule
Wear during sleep & Indicator value: 1,  worn during sleep or 0, otherwise \\
\midrule
Active hours & Number of hours with a step count > 500  \\
\midrule
Low active hours & Number of hours with a step count < 500  \\
\midrule
Mean hourly steps & Average of hourly step count \\
\midrule
Step count & Total step count per day \\
\midrule
Normalised step count & Rate of steps per minute when fitbit is worn $\frac{\textit{Step Count}}{\textit{Minutes Worn}}$ \\
\midrule
 Step active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day. A minute is considered active if it has a step count > 100 and is part of a sequence of at least $k$ consecutive minutes with step count > 100. Note that for $k=0$ there is no restriction on the number of consecutive minutes\\
\midrule
Step active minutes during waking period & Number of minutes with step > 100 from 6am to midnight \\
\midrule
Step coefficient of variation & Coefficient of variation of minute step counts $CoV = \frac{\sigma (\textit{Step count})}{\mu (\textit{Step count})} \cdot 100$ \\
 \midrule
 Normalised step count for periods of : &  Rate of steps per minute when the heartrate is: \\
 - moderate activity & - between 120 and 140. \\ 
 - vigorous activity & - above 140. \\
 - moderate and vigorous activity & - above 120. \\
 - non-moderate and non-vigorous activity & - below 120.\\
 \midrule
 Resting heartrate proxy & Mean of the 5 minutes with the lowest heartrate average \\
 \midrule
Heartrate active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day. A minute is considered active if it has heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes with heartrate > 120.\\
\midrule
Heartrate active minute during waking period & Number of minutes with heartrate > 120 from 6am to midnight \\
\midrule
Heartrate coefficient of variation & Coefficient of variation for heartrate minute averages $CoV = \frac{\sigma (\textit{Heartrate})}{\mu (\textit{Heartrate})} \cdot 100$ \\
 \midrule
 Number of switches above threshold $t$, for $t\in \{100,120,140\}$ & Number of times the heartrate goes above the threshold $t$ from below this threshold.\\ 
 \midrule
 Step and heartrate active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day. A minute is considered active if it has a step count > 100 and heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes which satisfy both conditions.
\midrule
Heartrate Hourly max, min and median & Maximum. minimum and median of average hourly heartrate. \\
\bottomrule
\end{longtable}

@@

\begin{longtable}{ n|d}
\toprule
\cmidrule(r){1-2}
\textbf{Feature Name} & \textbf{Feature description} \\
\midrule
Breath Count & Number of breaths the person has made though an ACT  \\ 
 ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
 \midrule
 Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
 Minimum Breath Amplitude & Similarly minimum, maximum and standard deviation of breath amplitude. \\
 Maximum Breath Amplitude & ~ \\
 Standard Deviation of Breath Amplitude & ~ \\
 \midrule
 Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment. \\ 
 Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
 Maximum Break Duration & breath duration in a treatment. \\
 Standard Deviation of Breath Duration & ~ \\
 \midrule
 Break Count & Number of breaks (between breaths) in a treatment. \\
 \midrule
 Mean Break Duration & Average duration (in seconds) of breaks between breaths. \\ 
 Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
 Maximum Break Duration & breaks duration in a treatment. \\
 Standard Deviation of Break Duration & ~ \\
 \midrule
 Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
 ~ & sets. Set is a group of breaths separated by huffing and coughing activity. \\
 \midrule
 Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
 ~ & Used to filter out "test" device sessions that are not part of a treatment. \\
 \midrule
 Pressures Readings Completeness Score & Data completeness score for the day \\
 ~ & If a treatment has less than 150 pressure readings it's completeness \\
 ~ & score is 0. \\
 \midrule
 Treatment Duration & Treatment duration (in minutes). \\
 \midrule
 Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
 ~ & given day). \\
 \midrule
 Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
 ~ & out of the prescribed amount of breaths. \\ 
 \midrule
 Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
 ~ & out of the prescribed number of treatments for this period. \\
 \midrule
 Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days \\
 ~ & (for a given date) out of the prescribed amount of breaths for this period. \\ 
 \midrule
 Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
 ~ & out of the prescribed number of treatments for this period. \\
 \midrule
 Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
 ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
 \midrule
 Cumulative Adherence Score & Cumulative adherence score to date. \\
 ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
 \midrule
 Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
 \midrule
 Skipped Days Count & Skipped days since the last treatment. \\
 \bottomrule
\end{longtable}

@@

\begin{table}[H]
  \caption{Steps clusters Metrics Kmeans}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Size & Average Silhouette width \\
    \midrule
    XS & 1099 & 0.95 \\
    S & 3942 & 0.34 \\
    M & 2737 & 0.27 \\
    L & 1982 & 0.24 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{table}[H]
  \caption{Steps clusters Metrics HC}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Cluster number & Cluster Size & Average Silhouette width \\
    \midrule
    1 & 4996 & 0.12 \\
    2 & 4056 & 0.11 \\
    3 & 677 & 0.18 \\
    4 & 31 & 0.41 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{table}[H]
  \caption{Heartrate clusters Metrics Kmeans}
  \label{hr_metrics}
  \centering
  \begin{tabular}{ |c|c|c|}
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Size & Average Silhouette width \\
    \midrule
    S & 1246 & 0.28 \\
    SM & 1640 & 0.2 \\
    M1 & 2308 & 0.21 \\
    M2 & 2584 & 0.26 \\
    L & 1982 & 0.33 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

Calinski-Harabasz index for the heartrate clusters is 135.51 .

\begin{table}[H]
  \caption{Heartrate clusters Metrics HC}
  \label{hr_metrics}
  \centering
  \begin{tabular}{ |c|c|c|}
    \toprule
    \cmidrule(r){1-2}
    Cluster number & Cluster Size & Average Silhouette width \\
    \midrule
    1 & 4221 & -0.07 \\
    2 & 4901 & -0.06 \\
    3 & 607 & -0.10 \\
    4 & 24 & 0.07 \\
    5 & 7 & -0.06 \\
    \bottomrule
    \end{tabular}
\end{table}
\section{Appendix : Cluster results and metrics for ACT treatments}

The tables below show the metrics for the ACT treatment clusters.
\subsection{K-Means Clustering}
\subsubsection{K-Means Metrics and Number of Clusters}
The table below shows comparison of Average Silhouette width and Calinski-Harabasz index different number of clusters

\begin{table}[H]
  \caption{Clusters Metrics Kmeans vs Number of Clusters}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Number of Clusters & Average Silhouette width & Calinski-Harabasz index \\
    \midrule
    3 & 2292 & 0.17 \\
    4 & 2208 & 0.18 \\
    5 & 1996 & 0.15 \\
    6 & 1830 & 0.15 \\
    \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Cluster interpretation}
Clustering with number clusters k=4 has combination of good performance (Average Silhouette width and Calinski-Harabasz index) and positive feedback from clinical team on cluster level granularity and clusters interpretability.

\begin{figure}
  \centering
  \caption{ACT clusters displayed on PCA and UMAP projections obtained using Kmeans clustering}
  \includegraphics[scale=0.5]{fig_ACT_4_clusters.png}
  \label{fig:figACT4clusters}
\end{figure}

\begin{table}[htb]
  \caption{ACT Clusters interpretation}
  \label{tab:act_interpret}
  \centering
  \begin{tabular}{ c|l}
    \toprule

  \textbf{Feature Name} & \textbf{Feature description} \\
  \midrule
    1 & Treatments with exceptionally big breath counts and biggest treatment adherence \\
  \midrule
    2 & Breath counts weekly adherence is on a lower side, treatments are short \\
  \midrule
    3 & Good breath counts and related breath weekly adherence, the breaths themselves have big standard  \\ 
    ~ &  deviation of duration \\
  \midrule
    4 & Very good breath counts and related breath weekly adherence,  breaths are mostly short and powerful \\ 
    ~ & (with biggest standard deviation for amplitude), shorter treatments  \\
  \bottomrule
    \end{tabular}
\end{table}
